\chapter{Modern Architecture Deep Dives}

The original GPT-2 architecture, introduced in 2019, established a template
that still shapes transformer design today: embedding lookups, multi-head
self-attention, LayerNorm, GELU feed-forward networks, and learned absolute
positional embeddings. Yet every component in that template has since been
reconsidered. Modern large language models---LLaMA, Mistral, Gemma, and
others---replace each piece with something better: LayerNorm becomes RMSNorm,
absolute positional embeddings become Rotary Positional Embeddings (RoPE),
GELU becomes SwiGLU, and standard Multi-Head Attention becomes Grouped Query
Attention (GQA).

Why? Three problems motivate every change in this chapter:

\begin{itemize}
  \item \textbf{Training stability.} LayerNorm's mean subtraction introduces
        a shift that can destabilize deep networks; RMSNorm removes it without
        hurting quality.
  \item \textbf{Context length.} Absolute positional embeddings give the model
        no way to generalize beyond the sequence lengths seen during training.
        RoPE encodes position through rotations, enabling extrapolation and
        making relative position a natural byproduct.
  \item \textbf{Inference efficiency.} During autoregressive generation, the
        KV cache grows linearly with sequence length and number of heads. GQA
        reduces the number of KV heads, cutting memory and bandwidth without
        significant quality loss.
\end{itemize}

This chapter dives into each replacement: what it changes, why it works, and
how to implement it. By the end, you will understand the full set of
architectural upgrades that separates a 2019-era GPT-2 from a 2024-era
open-weight LLM.

\section{RMSNorm}

LayerNorm normalizes each vector $\mathbf{x} \in \mathbb{R}^d$ by first
subtracting the mean and then dividing by the standard deviation:

\begin{align*}
  \mu &= \frac{1}{d}\sum_{j=1}^{d} x_j, \qquad
  \sigma^2 = \frac{1}{d}\sum_{j=1}^{d}(x_j - \mu)^2 \\[6pt]
  \hat{x}_i &= \frac{x_i - \mu}{\sqrt{\sigma^2 + \epsilon}}, \qquad
  y_i = \gamma\,\hat{x}_i + \beta
\end{align*}

RMSNorm (Root Mean Square Layer Normalization) simplifies this in two ways.
First, it removes the mean subtraction---it normalizes by the root mean square
alone. Second, it removes the bias $\beta$, keeping only a learnable gain
$\gamma$:

\begin{align*}
  \textbf{RMSNorm:} \quad \hat{x}_i = \frac{x_i}{\sqrt{\frac{1}{d}\sum_{j=1}^{d} x_j^2 + \epsilon}}, \quad y_i = \gamma\,\hat{x}_i
\end{align*}

\begin{keyconcept}
  \textbf{Why does removing mean subtraction work?} In transformers,
  normalizations are applied \emph{before} each sublayer (pre-norm). The
  residual path $x + f(\text{Norm}(x))$ already preserves the mean of $x$,
  so recentering inside the normalization is redundant. RMSNorm rescales
  the magnitude---the only part that grows unstably in deep
  networks---without shifting the center, and the learnable $\gamma$ provides
  all the expressivity needed.

  \textbf{Why no bias?} With no mean subtraction, the output is
  zero-centered by construction (the normalization itself centers around
  zero). A bias $\beta$ would merely re-introduce a shift that the
  normalization just removed, making it redundant.
\end{keyconcept}

\begin{keyconcept}
  \textbf{Float32 upcasting}: Both LayerNorm and RMSNorm must upcast their
  inputs to float32 before computing the reduction (mean, variance, or RMS).
  Low-precision types like bf16 and fp16 can overflow or lose precision in
  these reductions. After computing the normalization in float32, the result
  is cast back to the original dtype before the affine transform ($\gamma$),
  which stays in the original dtype.
\end{keyconcept}

\begin{pseudocode}[RMSNorm Forward Pass]
import torch

def rms_norm(x: torch.Tensor, weight: torch.Tensor, eps: float = 1e-6) -> torch.Tensor:
    # x: [batch, seq_len, embed_dim]
    # weight: [embed_dim]  (gamma)
    # Compute in float32 for numerical stability
    x_float = x.float()
    # Root mean square: sqrt(mean(x^2) + eps)
    rms = torch.sqrt(torch.mean(x_float ** 2, dim=-1, keepdim=True) + eps)
    # Normalize and cast back to original dtype
    x_normed = (x_float / rms).type_as(x)
    # Apply learnable gain (no bias)
    return weight * x_normed  # [batch, seq_len, embed_dim]
\end{pseudocode}

\section{RoPE (Rotary Positional Embeddings)}

GPT-2 uses \emph{learned absolute positional embeddings}---a lookup table that
adds a position-dependent vector to each token embedding. This has a fundamental
limitation: the model can only handle positions up to the maximum it was trained
on. If you train with 1{,}024 positions, position 1{,}025 has no embedding and
the model has no way to reason about it.

Rotary Positional Embeddings (RoPE) solve this by encoding position through
\emph{rotations} in the embedding space. Instead of adding a vector, RoPE
\emph{rotates} pairs of dimensions in the query and key tensors by an angle
that depends on the position. This gives two crucial properties:
\begin{enumerate}
  \item The dot product between a query at position $m$ and a key at position
        $n$ depends only on the \emph{relative distance} $m - n$, not the
        absolute positions. This is called \emph{relative position encoding}.
  \item Because rotations are defined by angle (a continuous value), RoPE can
        in principle extrapolate to positions beyond the training range---there
        is no hard cutoff.
\end{enumerate}

\subsection{The Rotation Operation}

RoPE operates on pairs of dimensions $(2i, 2i+1)$ of the query and key
vectors. For a vector $\mathbf{q} \in \mathbb{R}^d$ at position $m$, each
pair is rotated by an angle $m\theta_i$:

\[
  \begin{bmatrix} q_{2i}' \\ q_{2i+1}' \end{bmatrix}
  =
  \begin{bmatrix} \cos(m\theta_i) & -\sin(m\theta_i) \\ \sin(m\theta_i) & \cos(m\theta_i) \end{bmatrix}
  \begin{bmatrix} q_{2i} \\ q_{2i+1} \end{bmatrix}
\]

where $\theta_i = \text{base}^{-2i/d}$ (base $= 10{,}000$ by default). This
is a standard 2D rotation matrix applied independently to each pair of
coordinates. Different pairs use different rotation frequencies: low-index
pairs (near the start of the vector) rotate faster (large $\theta_i$), while
high-index pairs rotate slower (small $\theta_i$). The same rotation is
applied to both queries and keys, but with their respective positions.

\subsection{Why RoPE Gives Relative Position Encoding}

Consider what happens when you compute the dot product between a rotated
query at position $m$ and a rotated key at position $n$. For a single pair
of dimensions with index $i$:

\[
  q_{2i}' \cdot k_{2i}' + q_{2i+1}' \cdot k_{2i+1}'
\]

Substituting the rotated values:

\begin{align*}
  & \bigl(q_{2i}\cos m\theta_i - q_{2i+1}\sin m\theta_i\bigr)
    \bigl(k_{2i}\cos n\theta_i - k_{2i+1}\sin n\theta_i\bigr) \\
  & + \bigl(q_{2i}\sin m\theta_i + q_{2i+1}\cos m\theta_i\bigr)
    \bigl(k_{2i}\sin n\theta_i + k_{2i+1}\cos n\theta_i\bigr)
\end{align*}

Expanding and collecting terms, this simplifies to:

\[
  (q_{2i}k_{2i} + q_{2i+1}k_{2i+1})\cos\bigl((m-n)\theta_i\bigr)
  + (q_{2i}k_{2i+1} - q_{2i+1}k_{2i})\sin\bigl((m-n)\theta_i\bigr)
\]

The key insight: the result depends on $m$ and $n$ \textbf{only through their
difference} $(m - n)$. The absolute positions have cancelled out. This is a
direct consequence of the rotation group structure---applying rotation $m$ and
then the inverse of rotation $n$ yields rotation $m - n$.

\begin{keyconcept}
  \textbf{Critical properties of RoPE}:
  \begin{itemize}
    \item The dot product $q_m^\top k_n$ depends \emph{only on the relative
          distance} $(m - n)$, not on absolute positions $m$ and $n$
          individually.
    \item RoPE is applied \emph{only to Q and K, never to V}. Position
          information enters exclusively through attention scores.
    \item Different dimensions capture different position frequencies:
          low-index pairs are sensitive to fine-grained (nearby) position
          differences, while high-index pairs respond to long-range patterns.
  \end{itemize}
\end{keyconcept}

\subsection{Efficient Implementation: Rotate-and-Interleave}

Handling pairs of dimensions explicitly is cumbersome. The standard
implementation uses an equivalent formulation that operates on the full
vector at once:

\[
  \mathbf{q}' = \mathbf{q} \odot \cos(m\boldsymbol{\theta}) + \text{rotate\_half}(\mathbf{q}) \odot \sin(m\boldsymbol{\theta})
\]

where $\text{rotate\_half}$ splits the vector into pairs
$(q_{2i}, q_{2i+1})$, constructs $(-q_{2i+1}, q_{2i})$ for each pair, and
interleaves the result back into a single vector. This produces exactly the
same result as the explicit rotation but is far more efficient.

\begin{pseudocode}[Applying Rotary Positional Embeddings]
import torch

def apply_rotary_emb(
    x: torch.Tensor,
    cos: torch.Tensor,
    sin: torch.Tensor,
) -> torch.Tensor:
    # x:   [batch, num_heads, seq_len, head_dim]
    # cos: [batch, 1, seq_len, head_dim]  (broadcast over heads)
    # sin: [batch, 1, seq_len, head_dim]

    # Pair up dimensions: reshape to [..., head_dim/2, 2]
    x_pairs = x.float().reshape(*x.shape[:-1], -1, 2)
    # rotate_half: (q_2i, q_{2i+1}) -> (-q_{2i+1}, q_2i)
    x_rotated = torch.stack((-x_pairs[..., 1], x_pairs[..., 0]), dim=-1)
    x_rotated = x_rotated.reshape_as(x)
    # Apply rotation: q' = q * cos + rotate_half(q) * sin
    output = x * cos + x_rotated * sin
    return output.type_as(x)  # [batch, num_heads, seq_len, head_dim]
\end{pseudocode}

\begin{pseudocode}[Computing Rotary Frequencies]
import torch

def precompute_rope_freqs(
    head_dim: int,
    max_seq_len: int,
    base: float = 10000.0,
) -> tuple[torch.Tensor, torch.Tensor]:
    # theta_i = base^(-2i/d) for i = 0, 1, ..., head_dim/2 - 1
    freqs = 1.0 / (base ** (torch.arange(0, head_dim, 2).float() / head_dim))
    # Position indices: [0, 1, ..., max_seq_len - 1]
    positions = torch.arange(max_seq_len).float()
    # Angle for each (position, dimension_pair): m * theta_i
    angles = torch.outer(positions, freqs)  # [max_seq_len, head_dim/2]
    # Duplicate to match head_dim (each pair shares one angle)
    angles = torch.cat([angles, angles], dim=-1)  # [max_seq_len, head_dim]
    cos = angles.cos()  # [max_seq_len, head_dim]
    sin = angles.sin()  # [max_seq_len, head_dim]
    return cos, sin
\end{pseudocode}

\section{SwiGLU}

The feed-forward network (FFN) in the original transformer uses two linear
projections with a GELU activation in between:

\[
  \text{FFN}(x) = W_2\!\left(\text{GELU}(W_1 x + b_1)\right) + b_2
\]

This is a simple expansion--activation--contraction pipeline: the input is
projected up to a higher dimension (typically $4\times$ the embedding
dimension), passed through a smooth nonlinearity, and projected back down.

SwiGLU replaces this with a \emph{gated} architecture that uses three linear
projections and a multiplicative gate:

\[
  \text{FFN}(x) = W_{\text{down}}\!\left(W_{\text{gate}}(x) \odot \text{SiLU}(W_{\text{up}}(x))\right)
\]

where $\text{SiLU}(x) = x \cdot \sigma(x)$ is the Sigmoid Linear Unit (also
called ``swish''), and $\sigma(x) = \frac{1}{1 + e^{-x}}$ is the sigmoid
function.

\subsection{Understanding the Gating Mechanism}

The key innovation is the \emph{gate}: $W_{\text{gate}}(x)$ produces a value
that is multiplied element-wise with the SiLU-activated $W_{\text{up}}(x)$.
When $\text{SiLU}(W_{\text{up}}(x)) \approx 0$, the gate effectively
``closes''---that path is silenced. When $\text{SiLU}(W_{\text{up}}(x)) \approx
W_{\text{up}}(x)$ (for large positive inputs), the gate is ``open'' and
information flows through freely.

This is qualitatively different from GELU: the gate provides a data-dependent
on/off switch, allowing the network to selectively route information. GELU,
by contrast, is a single nonlinearity---it modulates but never fully blocks
a path (GELU's output is zero only at $-\infty$).

\subsection{SiLU vs.\ GELU}

Both functions are smooth, monotone approximations to ReLU, but they differ
in detail:

\begin{center}
\begin{tabular}{lcc}
\toprule
 & \textbf{GELU} & \textbf{SiLU} \\
\midrule
Formula & $x \cdot \Phi(x)$ & $x \cdot \sigma(x)$ \\
Approximation & Gaussian CDF & Logistic sigmoid \\
At $x = 0$ & $\approx 0$ & $= 0.5$ \\
As $x \to -\infty$ & $\to 0$ & $\to 0$ \\
As $x \to +\infty$ & $= x$ & $= x$ \\
\bottomrule
\end{tabular}
\end{center}

SiLU is cheaper to compute (sigmoid is a single exp, while GELU requires an
approximation involving tanh). In the SwiGLU context, this matters because
SiLU is applied to one of the two projections in every forward pass.

\subsection{Dimension Calculation and Parameter Budget}

The standard FFN expands from $d_{\text{model}}$ to $d_{\text{ff}}$ and back:

\[
  \text{Standard FFN}: \quad d_{\text{ff}} = 4\,d_{\text{model}}, \quad
  \text{params} = 2 \times d_{\text{model}} \times d_{\text{ff}} = 8\,d_{\text{model}}^2
\]

SwiGLU uses three projections. If we kept $d_{\text{ff}} = 4\,d_{\text{model}}$,
the parameter count would be:

\[
  \text{SwiGLU (naive)}: \quad 3 \times d_{\text{model}} \times d_{\text{ff}} = 12\,d_{\text{model}}^2
\]

That is $50\%$ more parameters than the standard FFN. To match the parameter
budget, LLaMA sets:

\[
  d_{\text{ff}} = \frac{2}{3} \times 4\,d_{\text{model}} = \frac{8}{3}\,d_{\text{model}}
\]

so the total becomes $3 \times d_{\text{model}} \times \frac{8}{3}\,d_{\text{model}}
= 8\,d_{\text{model}}^2$, matching the standard FFN. For a model with
$d_{\text{model}} = 4096$: $d_{\text{ff}} = 10923$ (rounded from $10922.67$).

\begin{keyconcept}
  \textbf{Key differences from standard FFN}:
  \begin{itemize}
    \item 3 linear layers instead of 2 ($W_{\text{gate}}$,
          $W_{\text{up}}$, $W_{\text{down}}$)
    \item The gate controls which features pass through: when
          $\text{SiLU}(W_{\text{up}}(x)) \approx 0$, the gate ``closes''
          and blocks information
    \item Despite having more parameters per layer, SwiGLU is
          empirically more parameter-efficient---it matches or exceeds
          GELU FFN quality with fewer total parameters
    \item The FFN hidden dimension is adjusted from $4d_{\text{model}}$ to
          $\frac{8}{3}d_{\text{model}}$ (for LLaMA) to keep the parameter
          count on par with standard FFN
  \end{itemize}
\end{keyconcept}

\begin{pseudocode}[SwiGLU Forward Pass]
import torch
import torch.nn as nn

class SwiGLUFFN(nn.Module):
    def __init__(self, embed_dim: int, hidden_dim: int):
        super().__init__()
        self.w_gate = nn.Linear(embed_dim, hidden_dim, bias=False)
        self.w_up   = nn.Linear(embed_dim, hidden_dim, bias=False)
        self.w_down  = nn.Linear(hidden_dim, embed_dim, bias=False)

    def forward(self, x: torch.Tensor) -> torch.Tensor:
        # x: [batch, seq_len, embed_dim]
        # Gated activation: gate * SiLU(up_proj)
        gate = self.w_gate(x)                  # [batch, seq_len, hidden_dim]
        up   = self.w_up(x)                    # [batch, seq_len, hidden_dim]
        activated = gate * torch.nn.functional.silu(up)
        # Project back down
        return self.w_down(activated)           # [batch, seq_len, embed_dim]
\end{pseudocode}

\section{\texorpdfstring{GQA}{GQA} (Grouped Query Attention)}

During autoregressive generation, the model must store and reload the key and
value tensors for all previous positions at every step. This \emph{KV cache}
is the primary memory bottleneck at inference time. For a model with $h$
attention heads, head dimension $d_k$, and sequence length $s$, the KV cache
for a single layer is:

\[
  \text{KV cache size} = 2 \times h \times s \times d_k \times \text{sizeof(dtype)}
\]

For LLaMA-2 70B ($h = 64$, $d_k = 128$, dtype = fp16) at sequence length
4{,}096, that is $2 \times 64 \times 4096 \times 128 \times 2 = 128$ MB per
layer, and $\sim$3.5 GB for all 80 layers. Grouped Query Attention (GQA)
reduces this by sharing KV heads across groups of query heads.

\subsection{MHA vs GQA vs MQA}

\begin{center}
\begin{tikzpicture}[
  node distance=0.3cm and 0.8cm,
  qnode/.style={rectangle, draw, fill=accent!30, minimum width=0.5cm, minimum height=0.5cm, font=\tiny},
  kvnode/.style={rectangle, draw, fill=highlight!30, minimum width=0.5cm, minimum height=0.5cm, font=\tiny},
  label/.style={font=\small\bfseries}
]
 
  \node[label] (mhlabel) {MHA};
  \foreach \i in {1,...,8} {
    \node[qnode, right=0.15cm of mhlabel] (q\i) at ({1.5+\i*0.4}, 0) {};
    \node[kvnode, below=0.2cm of q\i] {};
  }
  
  \node[label, right=1.2cm of q8] (gqlabel) {GQA};
  \foreach \i in {1,...,8} {
    \node[qnode] (gq\i) at ({7+\i*0.4}, 0) {};
  }
  \foreach \i in {1,2} { \node[kvnode] (gkv\i) at ({7.2+\i*1.6}, -0.7) {}; }
  \draw[-{Stealth}, gray] (gq1.south) -- (gkv1.north);
  \draw[-{Stealth}, gray] (gq2.south) -- (gkv1.north);
  \draw[-{Stealth}, gray] (gq3.south) -- (gkv1.north);
  \draw[-{Stealth}, gray] (gq4.south) -- (gkv1.north);
  \draw[-{Stealth}, gray] (gq5.south) -- (gkv2.north);
  \draw[-{Stealth}, gray] (gq6.south) -- (gkv2.north);
  \draw[-{Stealth}, gray] (gq7.south) -- (gkv2.north);
  \draw[-{Stealth}, gray] (gq8.south) -- (gkv2.north);
  
  \node[label, right=1.2cm of gq8] (mqlabel) {MQA};
  \foreach \i in {1,...,8} {
    \node[qnode] (mq\i) at ({12.5+\i*0.4}, 0) {};
  }
  \node[kvnode] (mkv) at ({14.1}, -0.7) {};
  \foreach \i in {1,...,8} {
    \draw[-{Stealth}, gray] (mq\i.south) -- (mkv.north);
  }
\end{tikzpicture}
\end{center}

\begin{center}
\begin{tabular}{lccc}
\toprule
 & \textbf{MHA} & \textbf{GQA} & \textbf{MQA} \\
\midrule
Query heads & $h$ & $h$ & $h$ \\
KV heads & $h$ & $h_{\text{kv}} < h$ & 1 \\
Group size & 1 & $h / h_{\text{kv}}$ & $h$ \\
KV cache savings & None & $\times\, h_{\text{kv}}/h$ & $\times\, 1/h$ \\
\bottomrule
\end{tabular}
\end{center}

In standard Multi-Head Attention (MHA), every query head has its own key and
value head. Multi-Query Attention (MQA) takes this to the opposite extreme:
a single KV head is shared by all query heads. GQA sits in between---query
heads are divided into $g$ groups, and each group shares one KV head. The
group size is $g = h / h_{\text{kv}}$. When $h_{\text{kv}} = h$, GQA reduces
to MHA; when $h_{\text{kv}} = 1$, it reduces to MQA.

Empirically, GQA with a moderate number of KV heads (e.g., 8 KV heads for 32
query heads in LLaMA-2) retains nearly all of MHA's quality while cutting KV
cache size and memory bandwidth by $4\times$. MQA can degrade quality,
especially on tasks requiring fine-grained attention distinctions.

\subsection{GQA Forward Pass}

The GQA forward pass has two stages: first, compute keys and values with the
smaller projection matrices (producing only $h_{\text{kv}}$ heads), then
expand them to match the number of query heads. The expansion is done by
\texttt{repeat\_interleave}, which repeats each KV head $g$ times along the
head dimension so that every query head in a group sees the same KV head.

\begin{pseudocode}[GQA Forward Pass]
import torch
import torch.nn as nn
import math

class GroupedQueryAttention(nn.Module):
    def __init__(self, embed_dim: int, num_heads: int, num_kv_heads: int):
        super().__init__()
        self.num_heads = num_heads
        self.num_kv_heads = num_kv_heads
        self.head_dim = embed_dim // num_heads
        self.group_size = num_heads // num_kv_heads

        self.w_q = nn.Linear(embed_dim, num_heads * self.head_dim, bias=False)
        self.w_k = nn.Linear(embed_dim, num_kv_heads * self.head_dim, bias=False)
        self.w_v = nn.Linear(embed_dim, num_kv_heads * self.head_dim, bias=False)
        self.w_o = nn.Linear(num_heads * self.head_dim, embed_dim, bias=False)

    def forward(self, x: torch.Tensor) -> torch.Tensor:
        batch, seq_len, embed_dim = x.shape

        # Project to queries, keys, values
        q = self.w_q(x)  # [batch, seq_len, num_heads * head_dim]
        k = self.w_k(x)  # [batch, seq_len, num_kv_heads * head_dim]
        v = self.w_v(x)  # [batch, seq_len, num_kv_heads * head_dim]

        # Reshape into heads
        q = q.view(batch, seq_len, self.num_heads, self.head_dim)
        k = k.view(batch, seq_len, self.num_kv_heads, self.head_dim)
        v = v.view(batch, seq_len, self.num_kv_heads, self.head_dim)

        # Transpose to [batch, heads, seq_len, head_dim]
        q = q.transpose(1, 2)
        k = k.transpose(1, 2)
        v = v.transpose(1, 2)

        # Expand KV heads: [batch, num_kv_heads, seq_len, head_dim]
        #           -> [batch, num_heads, seq_len, head_dim]
        k = k.repeat_interleave(self.group_size, dim=1)
        v = v.repeat_interleave(self.group_size, dim=1)

        # Apply rotary embeddings to Q and K (not V!)
        # q = apply_rotary_emb(q, cos, sin)
        # k = apply_rotary_emb(k, cos, sin)

        # Scaled dot-product attention
        scores = torch.matmul(q, k.transpose(-2, -1)) / math.sqrt(self.head_dim)
        # Apply causal mask (set future positions to -inf)
        attn_weights = torch.softmax(scores, dim=-1)
        attn_output = torch.matmul(attn_weights, v)

        # Reshape and project back
        attn_output = attn_output.transpose(1, 2).contiguous()
        attn_output = attn_output.view(batch, seq_len, embed_dim)
        return self.w_o(attn_output)  # [batch, seq_len, embed_dim]
\end{pseudocode}

\begin{keyconcept}
  \textbf{Why apply RoPE after KV expansion?} RoPE must be applied after the
  KV heads are expanded to match the query heads. If applied before expansion,
  each KV head would be rotated only once, and the \texttt{repeat\_interleave}
  would copy the same rotated head to all query heads in the group. While this
  is mathematically equivalent (since all query heads in a group share the same
  KV head and thus the same rotation), applying RoPE after expansion avoids
  subtle numerical issues and is the convention used in most implementations.

  \textbf{Degenerate cases}: When $h_{\text{kv}} = h$, GQA reduces to
  standard MHA with no expansion needed. When $h_{\text{kv}} = 1$, it
  produces Multi-Query Attention (MQA) where a single KV head is shared by
  all query heads.
\end{keyconcept}

\section*{Review Questions}

\begin{reviewquestion}[1]
  RMSNorm removes both mean subtraction and the bias parameter compared
  to LayerNorm. What is the intuition for why removing mean subtraction
  still works well?
\end{reviewquestion}

\begin{reviewquestion}[2]
  Both LayerNorm and RMSNorm upcast to float32 for the reduction
  computation. Explain why this is necessary and what could go wrong if
  the computation stayed in bf16 or fp16.
\end{reviewquestion}

\begin{reviewquestion}[3]
  Prove that with RoPE, the dot product $q_m^\top k_n$ depends only on
  the relative position $(m - n)$, not on absolute positions $m$ and $n$
  individually. (Hint: expand the 2D rotation for a single dimension pair
  and use the cosine angle subtraction formula.)
\end{reviewquestion}

\begin{reviewquestion}[4]
  In the RoPE frequency formula $\theta_i = \text{base}^{-2i/d}$, what
  happens when the base is very small (e.g., 100)? What about very large
  (e.g., 1{,}000{,}000)? How does the base affect the model's sensitivity
  to nearby vs.\ distant positions?
\end{reviewquestion}

\begin{reviewquestion}[5]
  RoPE is applied only to Q and K, never to V. Explain why position
  information should enter through attention scores rather than values.
\end{reviewquestion}

\begin{reviewquestion}[6]
  The \texttt{rotate\_half} operation transforms $(q_{2i}, q_{2i+1})$
  into $(-q_{2i+1}, q_{2i})$. Verify that the efficient formula
  $q' = q \odot \cos + \text{rotate\_half}(q) \odot \sin$ gives the
  same result as the explicit 2D rotation matrix.
\end{reviewquestion}

\begin{reviewquestion}[7]
  Compare the SiLU and GELU activation functions. Under what conditions
  is SiLU approximately equal to GELU? When do they diverge most?
\end{reviewquestion}

\begin{reviewquestion}[8]
  The SwiGLU FFN uses three linear projections instead of two. If
  $d_{\text{model}} = 4096$ and we want to match the standard FFN's
  parameter count ($8\,d_{\text{model}}^2$), what should the hidden
  dimension $d_{\text{ff}}$ be? Show the calculation.
\end{reviewquestion}

\begin{reviewquestion}[9]
  In SwiGLU, the gate $W_{\text{gate}}(x)$ is multiplied element-wise
  with $\text{SiLU}(W_{\text{up}}(x))$. Explain the role of the gate:
  what happens when $\text{SiLU}(W_{\text{up}}(x))$ is close to zero
  vs.\ close to identity?
\end{reviewquestion}

\begin{reviewquestion}[10]
  Consider a model with 32 query heads and 8 KV heads using GQA.
  During autoregressive generation, how much smaller is the KV cache
  compared to standard MHA? At what cost---if any---to model quality?
\end{reviewquestion}

\begin{reviewquestion}[11]
  Explain why \texttt{repeat\_interleave} is used in GQA rather than
  a simple \texttt{repeat}. What is the difference in how the heads
  are arranged along the head dimension?
\end{reviewquestion}

\begin{reviewquestion}[12]
  GQA with $h_{\text{kv}} = h$ is standard MHA and $h_{\text{kv}} = 1$
  is MQA. What are the tradeoffs between these extremes? Why might a
  middle value like $h_{\text{kv}} = h/4$ be preferred?
\end{reviewquestion}

\section*{Problems}

\begin{problem}[1]
  \textbf{Implement RMSNorm.} Write a \texttt{nn.Module} class
  \texttt{RMSNorm} that takes \texttt{embed\_dim} and \texttt{eps} as
  arguments. The module should have a single learnable parameter
  \texttt{weight} (initialized to ones) and should upcast to float32 for
  the RMS computation before casting back. Test it by verifying that
  (a) a zero input produces zero output and (b) scaling the input by a
  constant $c$ results in the same output (RMSNorm is scale-invariant).
\end{problem}

\begin{problem}[2]
  \textbf{Implement RoPE application to query and key tensors.} Write a
  function \texttt{apply\_rotary\_emb(q, k, cos, sin)} that applies rotary
  embeddings to the query and key tensors but \emph{not} the value tensor.
  Verify correctness by checking: (a) at position 0, $q$ and $k$ are
  unchanged; (b) for two positions $m$ and $n$, the dot product
  $q_m^\top k_n$ depends only on $m - n$.
\end{problem}

\begin{problem}[3]
  \textbf{Implement the SwiGLU FFN.} Write a \texttt{nn.Module} class
  \texttt{SwiGLUFFN} with three linear projections
  (\texttt{w\_gate}, \texttt{w\_up}, \texttt{w\_down}) and no biases.
  The module should compute
  $W_{\text{down}}(W_{\text{gate}}(x) \odot \text{SiLU}(W_{\text{up}}(x)))$.
  Set the hidden dimension to
  $\lfloor\frac{2}{3} \times 4 \times d_{\text{model}}\rfloor$ to match
  the standard FFN's parameter budget.
\end{problem}

\begin{problem}[4]
  \textbf{Implement the GQA forward pass.} Write a
  \texttt{GroupedQueryAttention} module that supports any number of KV
  heads between 1 and \texttt{num\_heads}. Use
  \texttt{repeat\_interleave} to expand KV heads after projection but
  before computing attention scores. Verify that
  (a) setting \texttt{num\_kv\_heads = num\_heads} produces the same
  output as standard MHA and (b) the KV cache size scales as
  $h_{\text{kv}} / h$ relative to MHA.
\end{problem}

\begin{problem}[5]
  \textbf{Benchmark KV cache sizes.} For a model with
  $d_{\text{model}} = 4096$, 32 heads, head dimension 128, and 32
  layers, compute the KV cache size (in MB) at sequence length 4096 for
  MHA, GQA with 8 KV heads, and MQA. Assume fp16 precision. How does
  the KV cache scale with sequence length for each variant?
\end{problem}