\chapter{The GPT Architecture}

Generative Pre-trained Transformers (GPT) are \emph{decoder-only} language models
that generate text one token at a time. Unlike the original Transformer, which
has both an encoder and a decoder for sequence-to-sequence tasks, GPT uses only
the decoder stack: the model reads the entire input sequence through
\emph{causal} (masked) self-attention, where each position can attend to itself
and all prior positions but not future ones. This autoregressive constraint makes
GPT a natural fit for language modeling---predict the next token given all
previous tokens.

The architecture is remarkably uniform: a stack of identical \emph{transformer
blocks}, each containing a multi-head self-attention layer and a position-wise
feed-forward network, wrapped in residual connections and layer normalization.
The same block is repeated $N$ times (12 for GPT-2 Small), with every block
using the same design but independent learned parameters. This chapter walks
through each component---embeddings, normalization, attention, the feed-forward
network, and how they assemble into a complete model---before covering the
strategies used at inference time to turn logits into coherent text.

\section{Token and Positional Embeddings}

A language model's input is a sequence of token IDs: integers that index into a
fixed vocabulary. The first step is to convert these discrete IDs into
continuous vector representations that the model can process. This is the role
of \emph{embeddings}.

\subsection{Token Embeddings}

A token embedding is a lookup table---a matrix $E \in \mathbb{R}^{V \times d}$
where $V$ is the vocabulary size and $d$ is the embedding dimension. Each row
of this matrix is a $d$-dimensional vector corresponding to one token in the
vocabulary. Given a token ID $t_i$, its embedding is simply:

\[
  \mathbf{e}_i = E[t_i] \in \mathbb{R}^{d}
\]

In PyTorch, this is implemented as \texttt{nn.Embedding(V, d)}, which
internally stores the matrix $E$ and performs the lookup via integer indexing.
The embedding vectors are initialized randomly (typically from
$\mathcal{N}(0, 0.02)$) and then learned during training. Over the course of
training, semantically similar tokens develop similar embedding vectors, since
they appear in similar contexts.

For GPT-2, $V = 50{,}257$ and $d = 768$ (for the Small variant), so the token
embedding matrix contains $50{,}257 \times 768 \approx 38.6\text{M}$
parameters---a significant fraction of the model's total parameters.

\subsection{Positional Embeddings}

Token embeddings alone are \emph{permutation-invariant}: the model cannot
distinguish ``the cat sat'' from ``sat the cat'' because each token is mapped
to the same vector regardless of where it appears. Without position information,
the model degenerates into a bag-of-words model and cannot learn word order.

GPT-2 uses \emph{learned} positional embeddings: another matrix
$P \in \mathbb{R}^{L \times d}$, where $L$ is the maximum sequence length
(1,024 for GPT-2). The positional embedding for position $i$ is:

\[
  \mathbf{p}_i = P[i] \in \mathbb{R}^{d}
\]

The final input representation for token $t_i$ at position $i$ is the
\emph{sum} of the two embeddings:

\[
  x_i = \text{Embed}_{\text{token}}(t_i) + \text{Embed}_{\text{pos}}(i)
\]

\begin{keyconcept}
  \textbf{Why addition instead of concatenation?} Concatenating the token and
  positional embeddings would require $2d$ dimensions, doubling the
  downstream computation. Addition works because both embeddings live in the
  same $\mathbb{R}^{d}$ space, and the model can learn to disentangle the
  position signal from the token signal. You can think of each embedding
  dimension as one ``register'' that stores a combination of token identity
  and position, rather than devoting separate dimensions to each.

  Alternatively, the original Transformer used sinusoidal positional encodings
  (fixed functions of position), while GPT-2 found that simply learning the
  position embeddings from data works well. Some later models (e.g., Rotary
  Position Embeddings / RoPE) encode position through multiplicative
  interactions rather than additive ones.
\end{keyconcept}

\begin{pseudocode}[Token and Positional Embeddings]
class GPT2Embeddings(nn.Module):
    def __init__(self, vocab_size: int, context_len: int, embed_dim: int):
        self.tok_embed = nn.Embedding(vocab_size, embed_dim)
        self.pos_embed = nn.Embedding(context_len, embed_dim)
        self.drop = nn.Dropout(0.1)

    def forward(self, token_ids: Tensor) -> Tensor:
        # token_ids: [batch, seq_len]
        seq_len = token_ids.size(1)
        positions = torch.arange(seq_len, device=token_ids.device)
        # positions: [seq_len]
        x = self.tok_embed(token_ids) + self.pos_embed(positions)
        # x: [batch, seq_len, embed_dim]
        return self.drop(x)
\end{pseudocode}

\section{Layer Normalization}

Deep neural networks suffer from \emph{internal covariate shift}: the
distribution of activations changes across layers, making training unstable.
Layer normalization (LayerNorm) addresses this by normalizing the activations
across the feature dimension of each token independently.

Given an input vector $\mathbf{x} \in \mathbb{R}^{d}$, LayerNorm computes:

\[
  \text{LayerNorm}(\mathbf{x}) = \boldsymbol{\gamma} \odot \frac{\mathbf{x} - \boldsymbol{\mu}}{\boldsymbol{\sigma} + \epsilon} + \boldsymbol{\beta}
\]

where $\boldsymbol{\mu}$ and $\boldsymbol{\sigma}$ are the mean and standard
deviation computed over the $d$ features of $\mathbf{x}$, $\epsilon$ is a
small constant ($10^{-5}$) for numerical stability, and
$\boldsymbol{\gamma}$ and $\boldsymbol{\beta}$ are learnable scale and shift
parameters (each of dimension $d$) that allow the model to undo the
normalization if needed.

\begin{keyconcept}
  \textbf{LayerNorm vs.\ BatchNorm.} Batch normalization computes statistics
  across the \emph{batch} dimension, making it sensitive to batch size and
  dependent on the other examples in the batch. LayerNorm computes statistics
  across the \emph{feature} dimension of each token independently. This makes
  it:
  \begin{itemize}
    \item Independent of batch size (works for batch size $= 1$)
    \item Identical at training and inference (no running statistics)
    \item Natural for variable-length sequences (each position normalized
          independently)
  \end{itemize}
  These properties make LayerNorm the standard choice for transformer models.
\end{keyconcept}

\section{Residual Connections}

Each sublayer in the transformer block is wrapped in a \emph{residual connection}
(also called a skip connection):

\[
  \text{out} = x + \text{Sublayer}(x)
\]

\begin{keyconcept}
  \textbf{Why residual connections?} In a deep network with many layers,
  gradients must flow backward through every layer. Without skip connections,
  each layer multiplies the gradient by its Jacobian, and the product of many
  small Jacobians can shrink gradients to near zero (\emph{vanishing gradients}).
  The residual connection provides a ``clean path'' for gradients:
  \[
    \frac{\partial}{\partial x}\bigl(x + f(x)\bigr) = I + \frac{\partial f}{\partial x}
  \]
  Even if $\frac{\partial f}{\partial x}$ is near zero, the $I$ (identity)
  term ensures that gradients flow through unchanged. This is why deep
  residual networks (ResNets, and transformers with $>$6 layers) can be
  trained effectively.

  In GPT-2, the residual stream---the path from input to output that skips
  every sublayer---carries information forward. Each sublayer reads from this
  stream, computes, and \emph{adds} its result back. This means later layers
  can always access the original token representations, not just the
  transformed versions.
\end{keyconcept}

\section{Pre-Norm vs Post-Norm}

The original Transformer used \emph{post-norm}: LayerNorm is applied
\emph{after} the sublayer and residual addition. GPT-2 uses \emph{pre-norm}:
LayerNorm is applied \emph{before} the sublayer.

\begin{center}
\begin{tabular}{lll}
\toprule
 & \textbf{Pre-Norm (GPT-2+)} & \textbf{Post-Norm (Original)} \\
\midrule
Formula & $x + f(\text{LN}(x))$ & $\text{LN}(x + f(x))$ \\
LN position & Before sublayer & After sublayer \\
Gradient flow & LN \emph{not} in residual path & LN \emph{in} residual path \\
Training stability & Stable without warmup & Requires warmup \\
Output magnitude & Controlled by LN & Can grow with depth \\
\bottomrule
\end{tabular}
\end{center}

\begin{keyconcept}
  In 12-layer networks, pre-norm produces gradients $\sim5\times$ larger
  at embedding layers compared to post-norm. This is because in
  post-norm, LayerNorm sits inside the residual path, and its gradients
  can shrink the signal flowing backward. Pre-norm keeps LayerNorm
  ``outside'' the residual path, preserving gradient magnitude.

  In practice, pre-norm also means the output of each sublayer is added to
  the residual stream \emph{unchanged}---its magnitude does not depend on the
  number of layers. In post-norm, each LayerNorm rescales the residual stream,
  and as depth increases, the variance of the output can grow or shrink
  unpredictably.
\end{keyconcept}

\section{The GELU Activation Function}

The feed-forward network in each transformer block uses the \textbf{Gaussian
Error Linear Unit} (GELU) activation function, defined as:

\[
  \text{GELU}(x) = x \cdot \Phi(x) = x \cdot P(Z \leq x), \quad Z \sim \mathcal{N}(0, 1)
\]

where $\Phi(x)$ is the CDF of the standard normal distribution. Intuitively,
GELU multiplies the input by the probability that a standard normal random
variable is less than or equal to that input. For large positive $x$, $\Phi(x)
\approx 1$ and GELU passes the value through. For large negative $x$,
$\Phi(x) \approx 0$ and GELU zeroes it out. For values near zero, GELU
provides a smooth transition rather than a hard cutoff.

In practice, GELU is approximated by the faster ``tanh approximation'':

\[
  \text{GELU}(x) \approx 0.5\,x\left(1 + \tanh\!\left[\sqrt{\frac{2}{\pi}}\left(x + 0.044715x^3\right)\right]\right)
\]

\begin{keyconcept}
  \textbf{GELU vs.\ ReLU.} Both are non-linear activations, but they differ in
  important ways:
  \begin{itemize}
    \item \textbf{ReLU} applies a hard threshold: $\text{ReLU}(x) = \max(0, x)$.
          It completely zeros out negative inputs and provides no gradient
          for them. This creates ``dead neurons'' that never activate.
    \item \textbf{GELU} provides a \emph{smooth}, probabilistic gating. Small
          negative values are not completely zeroed---they receive a small but
          non-zero output proportional to how likely they are to be
          ``important.'' This soft gating is differentiable everywhere, unlike
          ReLU which has a kink at $x = 0$.
    \item Empirically, GELU outperforms ReLU in transformer models. The smooth
          gradient landscape and state-dependent gating are hypothesized to
          help optimization and representation learning.
  \end{itemize}
  In PyTorch, GELU is available as \texttt{nn.GELU()} or
  \texttt{F.gelu()}.
\end{keyconcept}

\section{Feed-Forward Network}

Each transformer block contains a position-wise feed-forward network (FFN)
that operates independently on each position. ``Position-wise'' means the
same two linear transformations are applied to every token, but with different
input values---equivalent to a 1D convolution with kernel size 1.

The FFN has two linear layers with a GELU activation in between:

\[
  \text{FFN}(x) = W_2\!\left(\text{GELU}(W_1 x + b_1)\right) + b_2
\]

\begin{keyconcept}
  The FFN's hidden dimension is $4\times$ the embedding dimension
  (e.g., $768 \rightarrow 3072$ for GPT-2 Small). This expansion-then-contraction
  pattern serves two purposes:
  \begin{itemize}
    \item The \emph{expansion} to 4$d$ allows the model to learn richer
          intermediate representations in a higher-dimensional space, where
          features that are entangled in the lower-dimensional embedding can
          be separated.
    \item The \emph{contraction} back to $d$ compresses this richer
          representation, and the residual connection adds this compressed
          result back to the original stream.
  \end{itemize}
  Together, the FFN is where much of the model's ``knowledge'' is stored---it
  can be thought of as a key-value memory where the first layer stores patterns
  and the second layer retrieves and combines them.
\end{keyconcept}

\begin{pseudocode}[Feed-Forward Network]
class FeedForward(nn.Module):
    def __init__(self, embed_dim: int):
        self.net = nn.Sequential(
            nn.Linear(embed_dim, 4 * embed_dim),
            nn.GELU(),
            nn.Linear(4 * embed_dim, embed_dim),
        )
        self.drop = nn.Dropout(0.1)

    def forward(self, x: Tensor) -> Tensor:
        # x: [batch, seq_len, embed_dim]
        return self.drop(self.net(x))
        # output: [batch, seq_len, embed_dim]
\end{pseudocode}

\section{Transformer Block Assembly}

A single transformer block combines multi-head self-attention, a feed-forward
network, layer normalization, residual connections, and dropout. Using the
pre-norm architecture:

\begin{tikzpicture}[
  node distance=0.6cm,
  block/.style={rectangle, draw, rounded corners, minimum width=4.5cm,
                minimum height=0.7cm, font=\small},
  arrow/.style={-{Stealth[length=2.5mm]}, thick}
]
  \node[block, fill=accent!15] (input) {Input: $x$};
  \node[block, fill=accent!25, below=of input] (ln1) {LayerNorm};
  \node[block, fill=highlight!20, below=of ln1] (attn) {Multi-Head Attention};
  \node[block, fill=accent!10, below=of attn] (drop1) {Dropout + Residual};
  \node[block, fill=accent!25, below=of drop1] (ln2) {LayerNorm};
  \node[block, fill=highlight!20, below=of ln2] (ffn) {Feed-Forward (GELU)};
  \node[block, fill=accent!10, below=of ffn] (drop2) {Dropout + Residual};
  \node[block, fill=accent!15, below=of drop2] (output) {Output};

  \draw[arrow] (input) -- (ln1);
  \draw[arrow] (ln1) -- (attn);
  \draw[arrow] (attn) -- (drop1);
  \draw[arrow] (drop1) -- (ln2);
  \draw[arrow] (ln2) -- (ffn);
  \draw[arrow] (ffn) -- (drop2);
  \draw[arrow] (drop2) -- (output);
\end{tikzpicture}

The computation inside a single block is:

\begin{align*}
  x' &= x + \text{Dropout}\!\left(\text{MHA}\!\left(\text{LN}_1(x)\right)\right) \\
  x'' &= x' + \text{Dropout}\!\left(\text{FFN}\!\left(\text{LN}_2(x')\right)\right)
\end{align*}

Notice the pattern: each sublayer (attention or FFN) operates on the
\emph{normalized} input, and its output is added back to the
\emph{un-normalized} residual stream. This is the pre-norm scheme discussed
earlier: the residual stream always carries the original signal, and each
sublayer contributes an additive correction.

\begin{pseudocode}[Transformer Block Forward Pass]
class TransformerBlock(nn.Module):
    def __init__(self, embed_dim: int, num_heads: int, drop_rate: float = 0.1):
        self.ln1 = nn.LayerNorm(embed_dim)
        self.attn = MultiHeadAttention(embed_dim, num_heads, drop_rate)
        self.ln2 = nn.LayerNorm(embed_dim)
        self.ffn = FeedForward(embed_dim)
        self.drop1 = nn.Dropout(drop_rate)
        self.drop2 = nn.Dropout(drop_rate)

    def forward(self, x: Tensor) -> Tensor:
        # x: [batch, seq_len, embed_dim]
        # Pre-norm: normalize before sublayer, add residual after
        x = x + self.drop1(self.attn(self.ln1(x)))
        # x: [batch, seq_len, embed_dim]
        x = x + self.drop2(self.ffn(self.ln2(x)))
        # x: [batch, seq_len, embed_dim]
        return x
\end{pseudocode}

\begin{keyconcept}
  \textbf{Why dropout?} Dropout randomly zeros out elements during training with
  probability $p$ (typically 0.1 in GPT-2). It serves as a regularizer that
  prevents the model from becoming overly reliant on specific neurons. In the
  transformer block, dropout is applied \emph{after} each sublayer's output and
  \emph{before} it is added to the residual stream. At inference time, dropout
  is disabled (all neurons active), but outputs are scaled by $(1 - p)$ to
  maintain the expected activation magnitude.
\end{keyconcept}

\section{Full GPT-2 Architecture}

The complete GPT-2 model stacks $N$ transformer blocks on top of the
embeddings, adds a final LayerNorm, and projects to vocabulary-sized logits.
This output projection layer reuses the token embedding matrix (weight tying)
in some implementations, reducing the parameter count.

\begin{pseudocode}[GPT-2 Forward Pass]
class GPT2(nn.Module):
    def __init__(self, vocab_size: int = 50257,
                 context_len: int = 1024,
                 embed_dim: int = 768,
                 num_layers: int = 12,
                 num_heads: int = 12,
                 drop_rate: float = 0.1):
        self.embed = GPT2Embeddings(vocab_size, context_len, embed_dim)
        self.blocks = nn.ModuleList([
            TransformerBlock(embed_dim, num_heads, drop_rate)
            for _ in range(num_layers)
        ])
        self.ln_f = nn.LayerNorm(embed_dim)
        self.head = nn.Linear(embed_dim, vocab_size, bias=False)

    def forward(self, token_ids: Tensor) -> Tensor:
        # token_ids: [batch, seq_len]
        x = self.embed(token_ids)
        # x: [batch, seq_len, embed_dim]
        for block in self.blocks:
            x = block(x)
            # x: [batch, seq_len, embed_dim] after each block
        x = self.ln_f(x)
        # x: [batch, seq_len, embed_dim]
        logits = self.head(x)
        # logits: [batch, seq_len, vocab_size]
        return logits
\end{pseudocode}

\begin{center}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Embed Dim} & \textbf{Layers} & \textbf{Heads} & \textbf{Parameters} \\
\midrule
GPT-2 Small  & 768  & 12 & 12 & 124M \\
GPT-2 Medium & 1024 & 24 & 16 & 355M \\
GPT-2 Large  & 1280 & 36 & 20 & 774M \\
GPT-2 XL     & 1600 & 48 & 25 & 1.5B \\
\bottomrule
\end{tabular}
\end{center}

\begin{keyconcept}
  \textbf{Where are the parameters?} The bulk of GPT-2's parameters are in
  three places:
  \begin{itemize}
    \item \textbf{Token embeddings}: $V \times d = 50{,}257 \times 768 \approx 38.6\text{M}$
          parameters. This alone is $\sim$31\% of GPT-2 Small.
    \item \textbf{Attention layers}: Each head has four projections
          ($W_Q, W_K, W_V, W_O$), each $d \times d$, totaling $4d^2$ per
          block. Over 12 blocks, that's $4 \times 768^2 \times 12 \approx 28.3\text{M}$.
    \item \textbf{FFN layers}: Each block has two linear layers
          ($d \times 4d$ and $4d \times d$), totaling $8d^2$ per block.
          Over 12 blocks, that's $8 \times 768^2 \times 12 \approx 56.6\text{M}$.
  \end{itemize}
  The FFN layers account for roughly half the transformer blocks' parameters,
  which is why later models (like MoE architectures) often target the FFN for
  sparsification.
\end{keyconcept}

\section{Text Generation Strategies}

After the model produces logits of shape $[\text{batch}, \text{seq\_len},
\text{vocab\_size}]$, we need a strategy to select the next token from the
distribution over the vocabulary. The choice of strategy has a dramatic effect
on the quality, diversity, and coherence of generated text.

\subsection{Greedy Decoding}

The simplest strategy: at each step, pick the token with the highest
probability.

\begin{pseudocode}[Greedy Decoding]
def generate_greedy(model: nn.Module, token_ids: Tensor,
                    max_new_tokens: int) -> Tensor:
    # token_ids: [batch, seq_len]
    for _ in range(max_new_tokens):
        logits = model(token_ids)
        # logits: [batch, seq_len, vocab_size]
        next_logits = logits[:, -1, :]
        # next_logits: [batch, vocab_size]
        next_token = torch.argmax(next_logits, dim=-1, keepdim=True)
        # next_token: [batch, 1]
        token_ids = torch.cat([token_ids, next_token], dim=1)
        # token_ids: [batch, seq_len + 1]
    return token_ids
\end{pseudocode}

Greedy decoding is deterministic and fast, but it tends to produce repetitive,
boring text because it always chooses the most probable continuation, which in
natural language is often a safe, generic word. It can also get stuck in loops
(e.g., repeating the same phrase indefinitely).

\subsection{Temperature Sampling}

Temperature controls the ``sharpness'' of the probability distribution. After
computing logits, we divide by a temperature $T > 0$ before applying softmax:

\[
  P(w_i) = \frac{\exp(\text{logit}_i / T)}{\sum_j \exp(\text{logit}_j / T)}
\]

\begin{pseudocode}[Temperature Sampling]
def generate_with_temperature(model: nn.Module, token_ids: Tensor,
                              max_new_tokens: int, temperature: float) -> Tensor:
    # token_ids: [batch, seq_len]
    for _ in range(max_new_tokens):
        logits = model(token_ids)
        # logits: [batch, seq_len, vocab_size]
        next_logits = logits[:, -1, :] / temperature
        # next_logits: [batch, vocab_size]
        probs = torch.softmax(next_logits, dim=-1)
        # probs: [batch, vocab_size]
        next_token = torch.multinomial(probs, num_samples=1)
        # next_token: [batch, 1]
        token_ids = torch.cat([token_ids, next_token], dim=1)
    return token_ids
\end{pseudocode}

\begin{keyconcept}
  \textbf{Temperature effects:}
  \begin{itemize}
    \item $T \rightarrow 0$: softmax collapses to a point mass on the highest
          logit (greedy decoding).
    \item $T = 1$: use the model's original distribution (standard sampling).
    \item $T > 1$: the distribution becomes flatter, increasing randomness
          and creativity but risking incoherence.
    \item $T < 1$: the distribution becomes sharper, giving higher probability
          to top tokens. This produces more focused, deterministic text.
  \end{itemize}
  Typical values: $T \in [0.7, 1.0]$ for coherent but varied text.
\end{keyconcept}

\subsection{Top-$k$ Sampling}

Top-$k$ sampling restricts the candidate set to the $k$ tokens with the
highest probabilities. All tokens outside the top $k$ are set to zero
probability, and the remaining probabilities are renormalized.

\begin{pseudocode}[Top-k Sampling]
def generate_top_k(model: nn.Module, token_ids: Tensor,
                   max_new_tokens: int, temperature: float = 1.0,
                   k: int = 50) -> Tensor:
    # token_ids: [batch, seq_len]
    for _ in range(max_new_tokens):
        logits = model(token_ids)[:, -1, :]
        # logits: [batch, vocab_size]
        logits = logits / temperature
        # Zero out all tokens outside the top-k
        top_k_values, _ = torch.topk(logits, k)
        # top_k_values: [batch, k]
        threshold = top_k_values[:, -1:]  # k-th largest value
        # threshold: [batch, 1]
        logits = logits.masked_fill(logits < threshold, float('-inf'))
        # logits: [batch, vocab_size]  (non-top-k set to -inf)
        probs = torch.softmax(logits, dim=-1)
        # probs: [batch, vocab_size]
        next_token = torch.multinomial(probs, num_samples=1)
        # next_token: [batch, 1]
        token_ids = torch.cat([token_ids, next_token], dim=1)
    return token_ids
\end{pseudocode}

The advantage of top-$k$ is that it filters out unlikely tokens and prevents
the model from sampling absurd continuations. However, a fixed $k$ can be too
restrictive in some contexts (where the distribution should be broad) and too
permissive in others (where only one or two tokens are truly viable).

\subsection{Top-$p$ (Nucleus) Sampling}

Top-$p$ sampling (also called nucleus sampling) addresses the rigidity of
top-$k$ by adapting the candidate pool size. Instead of always keeping the top
$k$ tokens, it keeps the smallest set of tokens whose cumulative probability
exceeds $p$:

\begin{pseudocode}[Top-p (Nucleus) Sampling]
def generate_top_p(model: nn.Module, token_ids: Tensor,
                  max_new_tokens: int, temperature: float = 1.0,
                  p: float = 0.9) -> Tensor:
    # token_ids: [batch, seq_len]
    for _ in range(max_new_tokens):
        logits = model(token_ids)[:, -1, :] / temperature
        # logits: [batch, vocab_size]
        # Sort logits in descending order
        sorted_logits, sorted_indices = torch.sort(logits, descending=True)
        # sorted_logits: [batch, vocab_size]
        cumulative_probs = torch.cumsum(torch.softmax(sorted_logits, dim=-1), dim=-1)
        # cumulative_probs: [batch, vocab_size]
        # Remove tokens with cumulative probability above p (keep first token)
        sorted_mask = cumulative_probs - torch.softmax(sorted_logits, dim=-1) >= p
        # Zero out logits for removed tokens
        sorted_logits[sorted_mask] = float('-inf')
        # Unsort to restore original order
        logits = sorted_logits.scatter(1, sorted_indices, sorted_logits)
        # logits: [batch, vocab_size]  (non-nucleus set to -inf)
        probs = torch.softmax(logits, dim=-1)
        # probs: [batch, vocab_size]
        next_token = torch.multinomial(probs, num_samples=1)
        token_ids = torch.cat([token_ids, next_token], dim=1)
    return token_ids
\end{pseudocode}

\begin{keyconcept}
  Top-$p$ is more adaptive than top-$k$:
  \begin{itemize}
    \item When the model is \emph{confident} (one token dominates), top-$p$
          may keep only 1--2 tokens, keeping generation focused.
    \item When the model is \emph{uncertain} (probabilities are spread out),
          top-$p$ keeps many tokens, allowing creative continuations.
    \item Typical value: $p = 0.9$ means the model samples from tokens making
          up 90\% of the probability mass.
  \end{itemize}
  In practice, temperature and top-$p$ are often combined: temperature first
  adjusts the distribution's shape, then top-$p$ filters the candidates.
\end{keyconcept}

\section*{Review Questions}

\begin{reviewquestion}[1]
  Why does the GPT-2 architecture use pre-norm instead of post-norm?
  What specific problem does pre-norm solve?
\end{reviewquestion}

\begin{reviewquestion}[2]
  A TransformerBlock has two sublayers (attention and FFN), each wrapped
  in a residual connection and LayerNorm. Write out the full computation
  for both sublayers with pre-norm.
\end{reviewquestion}

\begin{reviewquestion}[3]
  Explain the difference between top-$k$ and top-$p$ sampling. In what
  situation would top-$p$ produce a different set of candidate tokens
  than top-$k$?
\end{reviewquestion}

\begin{reviewquestion}[4]
  Why are token and positional embeddings added rather than concatenated?
  What would be the disadvantage of concatenation?
\end{reviewquestion}

\begin{reviewquestion}[5]
  What is the purpose of the $4\times$ expansion in the feed-forward network?
  How many parameters does the FFN contribute per transformer block in
  GPT-2 Small?
\end{reviewquestion}

\begin{reviewquestion}[6]
  Explain why GELU can produce non-zero outputs for negative inputs while ReLU
  cannot. Why might this smooth gating be beneficial for representation
  learning in transformers?
\end{reviewquestion}

\begin{reviewquestion}[7]
  In the GPT-2 model, a final LayerNorm (\texttt{ln\_f}) is applied after all
  transformer blocks but before the output head. Why is this normalization
  necessary, and what would happen without it?
\end{reviewquestion}

\begin{reviewquestion}[8]
  When temperature $T$ approaches 0, what does sampling converge to? When
  $T$ approaches infinity, what happens to the probability distribution?
\end{reviewquestion}

\begin{reviewquestion}[9]
  In the residual connection $x' = x + f(\text{LN}(x))$, why is it important
  that $f$ operates on $\text{LN}(x)$ rather than $x$ directly? What property
  does this give the residual stream?
\end{reviewquestion}

\begin{reviewquestion}[10]
  GPT-2 uses learned positional embeddings instead of the sinusoidal
  positional encodings from the original Transformer. What are the advantages
  and disadvantages of each approach?
\end{reviewquestion}

\begin{reviewquestion}[11]
  In a model with 12 transformer blocks, each block adds two residual
  connections. How many ``skip paths'' exist from the input to the output?
  Why does this matter for gradient flow?
\end{reviewquestion}

\begin{reviewquestion}[12]
  The output head in GPT-2 (\texttt{nn.Linear(embed\_dim, vocab\_size)}) can
  optionally share weights with the token embedding matrix. What is the
  argument for weight tying, and how many parameters does it save for
  GPT-2 Small?
\end{reviewquestion}

\section*{Problems}

\begin{problem}[1]
  \textbf{Implement a TransformerBlock.} Write a PyTorch
  \texttt{nn.Module} that implements the pre-norm transformer block as described
  in this chapter. Your module should include:
  \begin{itemize}
    \item Two \texttt{LayerNorm} layers (\texttt{ln1}, \texttt{ln2})
    \item A \texttt{MultiHeadAttention} module
    \item A \texttt{FeedForward} module (with $4\times$ expansion and GELU)
    \item Dropout after each sublayer
    \item Residual connections around each sublayer
  \end{itemize}
  Verify that your block preserves the input shape, i.e., if the input is
  $[\text{batch}, \text{seq\_len}, \text{embed\_dim}]$, the output should be
  the same shape.
\end{problem}

\begin{problem}[2]
  \textbf{Implement text generation with temperature and top-$k$ sampling.}
  Starting from a trained model that outputs logits, write a
  \texttt{generate} function that:
  \begin{itemize}
    \item Takes a prompt as token IDs, a max number of new tokens, a
          temperature, and a $k$ value for top-$k$ filtering
    \item At each step, applies temperature scaling, then top-$k$
          filtering, then samples from the resulting distribution
    \item Concatenates each sampled token to the running sequence
    \item Stops when \texttt{max\_new\_tokens} is reached or an end-of-sequence
          token is produced
  \end{itemize}
  Generate text with temperature $\in \{0.5, 1.0, 1.5\}$ and $k = 50$.
  Observe how temperature affects the output.
\end{problem}

\begin{problem}[3]
  \textbf{Verify the parameter count.} Build a GPT-2 Small configuration
  ($\text{embed\_dim} = 768$, 12 layers, 12 heads, $\text{vocab\_size} = 50257$)
  and count the total number of parameters. Break down the count by component:
  token embeddings, positional embeddings, attention layers, FFN layers, and
  the output head. Compare your count to the published 124M figure. Where are
  the most parameters concentrated?
\end{problem}

\begin{problem}[4]
  \textbf{Compare greedy, temperature, and top-$p$ generation.} Load a
  pre-trained GPT-2 model and use the same prompt to generate text with each
  of the following strategies:
  \begin{itemize}
    \item Greedy (argmax)
    \item Temperature sampling with $T = 0.8$
    \item Top-$p$ sampling with $p = 0.9$ and $T = 0.8$
  \end{itemize}
  Generate 5 samples per strategy. For each strategy, report: (a) whether the
  output ever enters a repetition loop, (b) the diversity across 5 samples, and
  (c) your subjective assessment of coherence. What patterns do you notice?
\end{problem}

\begin{problem}[5]
  \textbf{Implement top-$p$ (nucleus) sampling.} Write a function that:
  \begin{itemize}
    \item Sorts logits in descending order
    \item Computes the cumulative softmax probabilities
    \item Identifies the smallest set of tokens whose cumulative probability
          exceeds $p$
    \item Sets logits for all tokens outside the nucleus to $-\infty$
    \item Samples from the resulting distribution
  \end{itemize}
  Compare the behavior of top-$p$ ($p = 0.9$) against top-$k$ ($k = 50$) on
  prompts where the model is very confident (e.g., ``The capital of France is'')
  and prompts where it is uncertain (e.g., ``The meaning of life is''). How
  does the candidate set size differ?
\end{problem}